---
title: "Groove Analysis"
author: "Janek Z"
format: html
theme: quartz
editor: visual
---

```{r, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
```
::: {.panel-tabset}
# Welcome

::: {.columns}
 
::: {.column width="55%"}

## Intro 

This repository was created for the course **Computational Musicology** at the University of Amsterdam.\
It contains an analysis of a selected corpus of music from my playlist *"I Don't Want This Groove to Ever End"*.

### The Playlist

The playlist *"I Don't Want This Groove to Ever End"*, named after its first track, is a collection of what I find groovy songs: **tracks that are heavily based on a looped or recurring musical phrase or rhythmic pattern**. Only songs of around 6 minutes or longer get added to this playlist, to allow the listener to immerse themselves in each one.

The tracks are mostly unique and vary greatly, and hearing a contrasting groove immediately after another can disrupt the listening experience. To prevent this, I have put a lot of thought into arranging their order.

### The Analysis

The goal of the conducted analysis will be to show how well some musical features can be used to determine the order of the playlist, based on the similarity of the tracks. The following musical featuers will be used:
- like Spotify track-level features, 
- chroma, timbre, and temporal features,
- and volume or loudness
:::
::: {.column width="5%"}
:::
::: {.column width="40%"}

<p style="margin-top: 0;">
You can view and listen to the playlist here:
</p>
<iframe data-testid="embed-iframe" style="border-radius:12px" src="https://open.spotify.com/embed/playlist/5R4pN6j7ogQeuaAVptvOnW?utm_source=generator&theme=0" width="100%" height="400" frameBorder="0" allowfullscreen="" allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture" loading="lazy"></iframe>
:::
:::

# Background

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(dplyr)
library(readr)
library(ggplot2)
```

```{r load playlist, include=FALSE}
playlist <- read_csv("~/Desktop/UvA/computational_musicology/groove_analysis/IDWTGTEE/I_Don't_Want_This_Groove_to_Ever_End.csv")

playlist <- playlist[1:(nrow(playlist) - 4), ]

playlist |>
  summarise(
    mean_speechiness = mean(Speechiness),
    mean_acousticness = mean(Acousticness),
    mean_liveness = mean(Liveness),
    sd_speechiness = sd(Speechiness),
    sd_acousticness = sd(Acousticness),
    sd_liveness = sd(Liveness),
    median_speechiness = median(Speechiness),
    median_acousticness = median(Acousticness),
    median_liveness = median(Liveness),
    mad_speechiness = mad(Speechiness),
    mad_acousticness = mad(Acousticness),
    mad_liveness = mad(Liveness)
  )
```

```{r number tracks, echo=FALSE}
playlist$track_number <- 1:nrow(playlist)
```
### Mix of features
After having loaded the playlist, I deleted the last four tracks from the data frame that will be analyzed, since they are not yet assigned a place in the playlist. I have then computed the transition distance across all features. This will allow me to check for spikes along the playlist length to see which moments or tracks have the least smooth transitions so that I can focus on those.\

```{r}
# compute transition distance
playlist <- playlist %>%
  mutate(
    d_energy = Energy - lag(Energy),
    d_tempo = Tempo - lag(Tempo),
    d_dance = Danceability - lag(Danceability)
  )

playlist <- playlist %>%
  mutate(
    transition_distance = sqrt(
      d_energy^2 + d_tempo^2 + d_dance^2
    )
  )

ggplot(playlist, aes(x = track_number, y = transition_distance)) +
  geom_line() +
  geom_point() +
  theme_minimal()
```

\
Note that the distances of subsequent points on the graph does not correspong to a bigger spike, but for each track the transition distance from the previous one is shown. This means that everywhere there is a spike, there is a large transition distance. <br> Also as the first song has none before it, it is omitted from the graph, we start with the second track.\
This graph shows that when we take a couple of the track-level features, namely: energy, tempo and danceability, the following transitions are between most different songs: 

```{r, echo=FALSE}
library(knitr)

top5_transitions <- playlist %>%
  mutate(previous_track = lag(`Track Name`)) %>%
  slice_max(order_by = transition_distance, n = 5) %>%
  select(previous_track, `Track Name`, track_number, transition_distance)

kable(top5_transitions, format = "html", digits = 2, caption = "Top 5 Playlist Transitions")
```
### Tempo
```{r, include=FALSE}
# check against randomness


# To make sure that my ordering of the track influenced the Spotify track level features of the playlist at all, I will compare the mean multivariate distance of both my playlist and the same songs from it ordered in a random way.


set.seed(123)

random_distances <- replicate(100, {
  shuffled <- playlist %>% sample_frac(1)
  
  d_energy <- diff(shuffled$Energy)
  d_valence <- diff(shuffled$Valence)
  d_tempo  <- diff(shuffled$Tempo)
  d_dance  <- diff(shuffled$Danceability)
  
  mean(sqrt(d_energy^2 + d_valence^2 + d_tempo^2 + d_dance^2))
})

mean_original <- mean(playlist$transition_distance, na.rm = TRUE)

mean_original
mean(random_distances)

# This shows that my ordering did indeed influence the Spotify track level features positively, as expected. The mean multivariate distance between tracks is lower in my playlist than in a randomly ordered one.

```


```{r tempo graph}
ggplot(playlist, aes(x = track_number, y = Tempo)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Tempo Across Playlist Order",
    x = "Playlist Position",
    y = "Tempo"
  ) +
  theme_minimal()
```

<br> This graph shows a few sudden spikes across the length of the playlist, although some of them can be eliminated. The first spike for example, which happens because track 7 has a tempo of 200, more than twice the tempo of the previous song. However, the song is very relaxed and I believe it can be perceived as half the speed. The same case can be seen on track 72 - Lose Yourself to Dance. That is why I will divide the tempos of these songs by two.

```{r}
#| results: hide
library(dplyr)

playlist <- playlist %>%
  mutate(Tempo = if_else(
    track_number %in% c(7, 62),
    Tempo / 2,
    Tempo
  ))


playlist %>%
  filter(track_number %in% c(7, 62)) %>%
  select(track_number, Tempo)
```
### Singular Features
Now that the bpm has been scaled, here are the graphs for tempo across tracks in the playlists, as well as danceability and energy.

```{r}
#| column: screen-inset-shaded
#| layout-nrow: 1
# tempo
ggplot(playlist, aes(x = track_number, y = Tempo)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Tempo Across Playlist Order",
    x = "Playlist Position",
    y = "Tempo"
  ) +
  theme_minimal()

# danceability
ggplot(playlist, aes(x = track_number, y = Danceability)) +
  geom_line() +
  geom_point() +
  labs(
    title = "danceability Across Playlist Order",
    x = "Playlist Position",
    y = "Danceability"
  ) +
  theme_minimal()

# energy
ggplot(playlist, aes(x = track_number, y = Energy)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Energy Across Playlist Order",
    x = "Playlist Position",
    y = "Energy"
  ) +
  theme_minimal()
```
<br>
These graphs show that seemingly the track-level features that best describe the genre across this playlist are tempo and danceability. Both parameters are most stable in the middle of the playlist between tracks 20-40 more or less. Danceability shows a slight rise and then fall, which would make sense according to how the playlist is structured, but it still has high differences between some neighbouring tracks.

```{r}
top5_tempo_transitions <- playlist %>%
  mutate(
    previous_track = lag(`Track Name`),
    tempo_diff = abs(Tempo - lag(Tempo))
  ) %>%
  slice_max(order_by = tempo_diff, n = 5) %>%
  select(previous_track, `Track Name`, track_number, tempo_diff)

kable(top5_tempo_transitions, format = "html", digits = 2, caption = "Top 5 Tempo Transitions")
```
```{r}
top5_danceability_transitions <- playlist %>%
  mutate(
    previous_track = lag(`Track Name`),
    danceability_diff = abs(Danceability- lag(Danceability))
  ) %>%
  slice_max(order_by = danceability_diff, n = 5) %>%
  select(previous_track, `Track Name`, track_number, danceability_diff)

kable(top5_danceability_transitions, format = "html", digits = 2, caption = "Top 5 Danceability Transitions")
```
### Chosen Tracks
Based on this info, I decided to compare the following neighbouring songs:\
- *Find Your State of Mind* and *I'll Be With You* (tracks 6 and 7), because the songs differ very much in the allround transition score, but not so much in the individual classifications, so I wonder how they differ from eachoter\
- *Goldrush* and *Words 2b Heard Meets Planetary Funk Alert* (tracks 69 and 70), because they differ a lot on many aspects and I wonder if there are some features that show that I have placed them next to each other correctly. 

# Clustering
```{r kmeans scaling}
features <- playlist %>%
  select(Energy, Valence, Danceability, Tempo, Acousticness)

features_scaled <- scale(features)
```

```{r kmeans3}
# 3 clusters
set.seed(12)

kmeans_result <- kmeans(features_scaled, centers = 3, nstart = 25)

playlist$cluster <- kmeans_result$cluster

ggplot(playlist, aes(x = track_number, y = cluster)) +
  geom_point() +
  geom_line() +
  theme_minimal()
```

<br>


```{r kmeans5}
# 5 clusters
set.seed(13)

kmeans_result <- kmeans(features_scaled, centers = 5, nstart = 25)

playlist$cluster <- kmeans_result$cluster

ggplot(playlist, aes(x = track_number, y = cluster)) +
  geom_point() +
  geom_line() +
  theme_minimal()
```

<br>

```{r}
# check optimal number of clusters
set.seed(123)

wss <- sapply(seq(1, 11, by = 2), function(k) {
  kmeans(features_scaled, centers = k, nstart = 25)$tot.withinss
})

elbow_df <- data.frame(
  k = seq(1, 11, by = 2),
  wss = wss
)

ggplot(elbow_df, aes(x = k, y = wss)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Elbow Method for Optimal k",
    x = "Number of Clusters (k)",
    y = "Within-Cluster Sum of Squares"
  ) +
  theme_minimal()
```

<br>
<!-- ### Select a Feature to View -->

<!-- ```{r, echo=FALSE} -->
<!-- # Functions to generate plot, table, and conclusion based on selected feature -->

<!-- q_feature_plot <- function(feature) { -->
<!--   if (feature == "tempo") { -->
<!--     ggplot(playlist, aes(x = track_number, y = Tempo)) + -->
<!--       geom_line() + geom_point() + -->
<!--       labs(title="Tempo Across Playlist Order", x="Playlist Position", y="Tempo") + -->
<!--       theme_minimal() -->
<!--   } else if (feature == "liveness") { -->
<!--     ggplot(playlist, aes(x = track_number, y = Liveness)) + -->
<!--       geom_line() + geom_point() + -->
<!--       labs(title="Liveness Across Playlist Order", x="Playlist Position", y="Liveness") + -->
<!--       theme_minimal() -->
<!--   } else if (feature == "energy") { -->
<!--     ggplot(playlist, aes(x = track_number, y = Energy)) + -->
<!--       geom_line() + geom_point() + -->
<!--       labs(title="Energy Across Playlist Order", x="Playlist Position", y="Energy") + -->
<!--       theme_minimal() -->
<!--   } -->
<!-- } -->

<!-- q_feature_table <- function(feature) { -->
<!--   if (feature == "tempo") top5_tempo_transitions -->
<!--   else if (feature == "Liveness") top5_liveness_transitions -->
<!--   else top5_energy_transitions -->
<!-- } -->

<!-- q_feature_conclusion <- function(feature) { -->
<!--   if (feature == "tempo") "Conclusion: This shows spikes at tracks 7 and 62, etc." -->
<!--   else if (feature == "Liveness") "Conclusion: Peaks indicate live-feel transitions." -->
<!--   else "Conclusion: Large jumps in energy show strong contrast." -->
<!-- } -->
<!-- ``` -->

<!-- ```{r, echo=FALSE} -->
<!-- # Display plot, table, and conclusion for the selected feature -->
<!-- q_feature_plot(feature) -->
<!-- q_feature_table(feature) -->
<!-- cat(q_feature_conclusion(feature)) -->
<!-- ``` -->



<!-- # Background -->

<!-- ::: {.card title="Assesing the Order"} -->
<!-- ```{r child="Untitled.Rmd", echo=FALSE, results='asis'} -->
<!-- knitr::knit_child('Untitled.Rmd', envir = list(feature = feature)) -->
<!-- ``` -->
<!-- ::: -->

# Chroma features [pitch]

I have made these two self similarity matrices, based on chroma features in:
